Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
fixed lemma renaming
Updated

Note that in the following we will make wide use of the following properties.
\begin{assumption}\label{asm:unbiasedness}
        We consider an estimate $\wh{\nabla}_N J(\vtheta)$ as in Eq.~\ref{E:policygradient.estimate} such that
\begin{enumerate}
        \item \textit{On-policy Unbiased Estimator.} 
                \[
                        \EVV[\tau_i \sim \pi(\cdot|\vtheta)]{\wh{\nabla}_N J(\vtheta)} = \EVV[\tau_i \sim \pi(\cdot|\vtheta)]{\frac{1}{N} \sum_{i=0}^N g(\tau_i|\vtheta)} = \nabla J(\vtheta)
                \]
        \item \textit{Off-policy Unbiased Estimator.}
                \[
        \EVV[\tau_i \sim \Delta(\cdot|\pi^B)]{\frac{1}{N} \sum_{i=0}^N \omega(\tau_i|\pi^B,\vtheta) g(\tau_i|\vtheta) } 
        = \nabla J(\vtheta) 
                \]
\end{enumerate}
\end{assumption}
Note that these assumptions are verified by REINFORCE and G(PO)MDP.

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.

@@

\subsection*{Basic Properties of SVRPG}
\properties
\begin{proof}
	Claim (\ref{eq:unbiased}) is from the unbiasedness of the estimators:
	\begin{align*}
	\EVV[]{\gradBlack{\vtheta}} &= \EVV[]{\gradApp{\wt{\vtheta}}{N}}  + \EVV[]{\gradApp{\vtheta}{B}} \\
	&\qquad- \EVV[]{\frac{1}{B}\sum_{i=0}^{B-1}\omega(\tau_i|\vtheta, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} \\
	&= \gradJ{\wt{\vtheta}} + \gradJ{\vtheta} - \gradJ{\wt{\vtheta}} = \gradJ{\vtheta}.
	\end{align*}
	As for claim (\ref{eq:zerovar}), note that as $\vtheta\to\vtheta^*$, also $\wt{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
	\begin{align*}
	\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[g(\tau|\vtheta^*)\right. \\
	&\left.- \cancel{\omega(\tau|\vtheta^*,\vtheta^*)}g(\tau|\vtheta^*)\right]
	= \Var\left[\gradApp{\vtheta^*}{N}\right].
	\end{align*}
	Note that it is essential that the trajectories used in the second and the third term are the same for the variance to vanish.
\end{proof}

@@

%New Lemma on Variance
\begin{restatable}[]{lemma}{varineq}\label{lemma:varineq}
For any random vector X, the variance (as defined in Definition \ref{def:var}), can be bounded as follows:
\[
	\Vars{X} \leq \Es{\norm[]{X}^2}.
\]
\end{restatable}
\begin{proof}
By using basic properties of expected value and scalar variance:
\begin{align*}
	\Vars{X} = \Es{\norm[]{X-\Es{X}}^2} &= \Es{\sum_{i=1}^{\dim(X)}\left(X_i-\Es{X_i}\right)^2} = \sum_{i=1}^{\dim(X)}\Es{\left(X_i-\Es{X_i}\right)^2} \\
	&\leq \sum_{i=1}^{\dim(X)}\Es{X_i^2} = \Es{\sum_{i=1}^{\dim(X)}X_i^2} = \Es{\norm[]{X}^2}.
\end{align*}
\end{proof}

@@

	&\leq \Ets[t-1]{\norm[2]{\gradJ{\vtheta_t^{s+1}}}^2} 

@@

    +\Ets{\norm[2]{

@@

			\Ets{\norm[2]{

@@

where (\ref{neweq:2}) is from (\ref{neweq:1}), (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from Lemma \ref{lemma:varineq}, (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
In this appendix we provide more details on the practical variants of our algorithm.
\subsection{Learning Rates}
\begin{algorithm}[h]
	\begin{algorithmic}
		\STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
		\STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
		\STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
		\STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
		\STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
		\STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
		\STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
	\end{algorithmic}
	\caption{
		\label{A:adam}
		Adam}
\end{algorithm}
Let us give a deeper insight on the two different learning rate schedules used by our algorithm. We report pseudo-code of the original ADAM in Algorithm \ref{A:adam}. As mentioned, we use two distinct instances of ADAM: one related to the snapshots, and one to the sub-iterations. In this way the ADAM associated to the snapshots will take into account only the history of gradient moments at the snapshots. The same reasoning holds for the sub-iterations. This, of course, allows our algorithm to take into account the different variabilities in a more accurate manner.
\subsection{Baseline}
For all batch gradient-based algorithms, we use the same time-varying feature encoding for the linear baseline:
\begin{align*}
\psi_{s,t}=concat(s, s \odot s, 0.01t, (0.01t)^2, (0.01t)^3,1)
\end{align*}
where $s$ is the state vector and $\odot$ is the element-wise product. In the case of our algorithm the model representing the baseline is fitted every snapshot so we have:\\
\begin{align*}
        \nabla J(\pi^T) =\frac{1}{N} \sum_{n=1}^{N}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(y_k^n)\right) (\gamma^h \Reward(y_h^n)-b_{sn})
          + \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\\ - \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\omega(z_{0:h}|\pi^B,\pi^T)
\end{align*}
Where $b_{sn}$ is the prediction given by the baseline at the snapshot, whereas $b_{si}$ is the prediction given by the baseline at a given subiteration. 
\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

>>>>

Note that in the following we will make wide use of the following properties.
\begin{assumption}\label{asm:unbiasedness}
        We consider an estimate $\wh{\nabla}_N J(\vtheta)$ as in Eq.~\ref{E:policygradient.estimate} such that
\begin{enumerate}
        \item \textit{On-policy Unbiased Estimator.} 
                \[
                        \EVV[\tau_i \sim p(\cdot|\vtheta)]{\wh{\nabla}_N J(\vtheta)} = \EVV[\tau_i \sim p(\cdot|\vtheta)]{\frac{1}{N} \sum_{i=0}^N g(\tau_i|\vtheta)} = \nabla J(\vtheta)
                \]
        \item \textit{Off-policy Unbiased Estimator.}
                \[
        \EVV[\tau_i \sim p(\cdot|\pi^B)]{\frac{1}{N} \sum_{i=0}^N \omega(\tau_i|\pi^B,\vtheta) g(\tau_i|\vtheta) } 
        = \nabla J(\vtheta) 
                \]
\end{enumerate}
\end{assumption}
Note that these assumptions are verified by REINFORCE and G(PO)MDP.

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.
In the rest of the proof we will always consider the $\ell_2$-norm when not explicitly stated.

@@

\subsection*{Basic Properties of SVRPG}
\properties*
\begin{proof}
	Claim (\ref{eq:unbiased}) is from the unbiasedness of the estimators:
	\begin{align*}
	\EVV[]{\gradBlack{\vtheta}} &= \EVV[]{\gradApp{\wt{\vtheta}}{N}}  + \EVV[]{\gradApp{\vtheta}{B}} \\
	&\qquad- \EVV[]{\frac{1}{B}\sum_{i=0}^{B-1}\omega(\tau_i|\vtheta, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})} \\
	&= \gradJ{\wt{\vtheta}} + \gradJ{\vtheta} - \gradJ{\wt{\vtheta}} = \gradJ{\vtheta}.
	\end{align*}
	As for claim (\ref{eq:zerovar}), note that as $\vtheta\to\vtheta^*$, also $\wt{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
	\begin{align*}
	\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[g(\tau|\vtheta^*)\right. \\
	&\left.- \cancel{\omega(\tau|\vtheta^*,\vtheta^*)}g(\tau|\vtheta^*)\right]
	= \Var\left[\gradApp{\vtheta^*}{N}\right].
	\end{align*}
	Note that it is essential that the trajectories used in the second and the third term are the same for the variance to vanish.
\end{proof}

@@

%New Lemma on Variance
The following lemma is trivial but we report for sake of clarity.

\begin{restatable}[Variance is upper-bounded by second moment.]{lemma}{varineq}\label{lemma:varineq}
For any random vector X, the variance (as defined in Definition \ref{def:var}), can be bounded as follows:
\[
	\Vars{X} \leq \Es{\norm[]{X}^2}.
\]
\end{restatable}
\begin{proof}
        By using Equation~\eqref{neweq:1} and the definition of variance of a scalar random variable:
\begin{align*}
	\Vars{X} = \Es{\norm[]{X-\Es{X}}^2} &= \Es{\sum_{i=1}^{\dim(X)}\left(X_i-\Es{X_i}\right)^2} = \sum_{i=1}^{\dim(X)}\Es{\left(X_i-\Es{X_i}\right)^2} \\
                                     &=\sum_{i=1}^{\dim(X)} \Es{X_i^2} - \left(\Es{X_i} \right)^2 \leq \sum_{i=1}^{\dim(X)}\Es{X_i^2}
%    = \Es{\sum_{i=1}^{\dim(X)}X_i^2} 
                                        = \Es{\norm[]{X}^2}.
\end{align*}
\end{proof}

@@

@@

    +\Ets{\norm[]{

@@

			\Ets{\norm[]{

@@

where (\ref{neweq:2}) is from (\ref{neweq:1}), (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$,~\ref{eq:2} is from Lemma~\ref{lemma:varineq}, (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

\section{Practical SVRPG versions}\label{app:practicalsvrpg}
In this appendix we provide more details on the practical variants of our algorithm.
\subsection{Learning Rates}
\begin{algorithm}[h]
	\begin{algorithmic}
		\STATE \textbf{Input:} A gradient estimate $g_t$ and parameters $\beta_1$, $\beta_2$, $\epsilon$ and $\alpha$.
		\STATE $m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t$
		\STATE $v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \circ g_t$ ($\circ$ is the Hadamard (component-wise) product)
		\STATE $\hat{m}_t = \dfrac{m_t}{1 - \beta^t_1}$
		\STATE $\hat{v}_t = \dfrac{v_t}{1 - \beta^t_2}$
		\STATE $\Delta(g_t) = \dfrac{\alpha}{\sqrt{\hat{v}_t} + \epsilon} \hat{m}_t$
		\STATE \textbf{Return:} The increment $\Delta(g_t)$  of the parameters.
	\end{algorithmic}
	\caption{
		\label{A:adam}
		Adam}
\end{algorithm}
Let us give a deeper insight on the two different learning rate schedules used by our algorithm. We report pseudo-code of the original ADAM in Algorithm \ref{A:adam}. As mentioned, we use two distinct instances of ADAM: one related to the snapshots, and one to the sub-iterations. In this way the ADAM associated to the snapshots will take into account only the history of gradient moments at the snapshots. The same reasoning holds for the sub-iterations. This, of course, allows our algorithm to take into account the different variabilities in a more accurate manner.
\subsection{Baseline}
For all batch gradient-based algorithms, we use the same time-varying feature encoding for the linear baseline:
\begin{align*}
\psi_{s,t}=concat(s, s \odot s, 0.01t, (0.01t)^2, (0.01t)^3,1)
\end{align*}
where $s$ is the state vector and $\odot$ is the element-wise product. In the case of our algorithm the model representing the baseline is fitted every snapshot so we have:\\
\begin{align*}
        \nabla J(\pi^T) =\frac{1}{N} \sum_{n=1}^{N}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(y_k^n)\right) (\gamma^h \Reward(y_h^n)-b_{sn})
          + \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\\ - \frac{1}{B} \sum_{n=1}^{B}
                \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k^n)\right) (\gamma^h \Reward(z_h^n) - b_{si})\omega(z_{0:h}|\pi^B,\pi^T)
\end{align*}
Where $b_{sn}$ is the prediction given by the baseline at the snapshot, whereas $b_{si}$ is the prediction given by the baseline at a given subiteration. 
\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
    \item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2]{a}^2$. The episodes terminates when the number of time steps T is greater than 500.
    \item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2]{a}^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}
